---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
library(Rdisop)
```

# Rdisop

<!-- badges: start -->
[![R-CMD-check](https://github.com/janlisec/Rdisop/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/janlisec/Rdisop/actions/workflows/R-CMD-check.yaml)
<!-- [![test-coverage](https://github.com/janlisec/Rdisop/actions/workflows/test-coverage.yaml/badge.svg)](https://github.com/janlisec/Rdisop/actions/workflows/test-coverage.yaml) -->
[![DOI](https://img.shields.io/badge/doi-10.1093/bioinformatics/btm631-yellow.svg)](https://doi.org/10.1093/bioinformatics/btm631)
<!-- badges: end -->

In high resolution mass spectrometry (HR-MS), the measured masses can be 
decomposed into potential element combinations (chemical sum formulas). 
Where additional mass/intensity information of respective isotopic peaks is
available, decomposition can take this information into account to better
rank the potential candidate sum formulas. To compare measured mass/intensity
information with the theoretical distribution of candidate sum formulas, the
latter needs to be calculated.
This package implements fast algorithms to address both tasks, the 
calculation of isotopic distributions for arbitrary sum formulas (assuming a
HR-MS resolution of roughly 30,000), and the ranked list of sum formulas
fitting an observed peak or isotopic peak set.

## Installation

You can install the development version of **Rdisop** using:

``` r
# install.packages("devtools")
devtools::install_github("sneumann/Rdisop")
```

## Example

The user can use built in sets of chemical elements or define such sets 
specifically and get informations on a specified sum formula, *i.e.* the
isotopic distribution which would be observed in HR-MS.

```{r exmpl2}
ele <- initializeElements(c("C","H","N","O","Mg"))
chlorophyll <- getMolecule("C55H72MgN4O5H", elements = ele)
getIsotope(chlorophyll, 1:4)
```

For individual masses, potential molecules can be calculated.

```{r exmpl3}
decomposeMass(mass = 46.042, ppm = 20, maxisotopes = 4)
```

This decomposition can also be performed for measurements containing several 
isotopes.

```{r exmpl5}
# glutamic acid (C5H9NO4)
mol <- decomposeIsotopes(masses = c(147.053, 148.056), intensities = c(93, 5.8))
data.frame(getFormula(mol), getScore(mol), getValid(mol))
```
